Making Sequential Programs Deterministic

### 2019-06-05 conclusion.tex
Use external bibliography when possible

We designed a type system that helps programmers express determinism specifications
in sequential programs. 
We implemented our type system in Java, applied it to real world 
software, and found it to be effective in finding previously unknown determinism bugs. 
Our tool \theDeterminismChecker found errors that the
Randoop developers had missed, despite spending
% over 100 hours
extensive effort on the
problem of nondeterminism. We compared our tool with existing state of the art flaky test detectors
and found that \theDeterminismChecker found all the bugs arising due to nondeterminism
in sequential programs that were used for evaluation in NonDex~\cite{nondex} and DeFlaker~\cite{deflaker}.
To the best of our knowledge, ours is the
first compile-time verification approach addressing the problem of
nondeterminism in sequential programs.

>>>>

We designed a type system that helps programmers express determinism specifications
in sequential programs. 
We implemented our type system in Java, applied it to real world 
software, and found it to be effective in finding previously unknown determinism bugs. 
Our tool \theDeterminismChecker found errors that the
Randoop developers had missed, despite spending
% over 100 hours
extensive effort on the
problem of nondeterminism. We compared our tool with existing state of the art flaky test detectors
and found that \theDeterminismChecker found all the bugs arising due to nondeterminism
in sequential programs that were used for evaluation in NonDex~\cite{ShiGLM2016} and DeFlaker~\cite{BellLHEYM2018}.
To the best of our knowledge, ours is the
first compile-time verification approach addressing the problem of
nondeterminism in sequential programs.

### 2019-06-05 experiments.tex
final edits
Use external bibliography when possible

The state of the art in flaky test detection is NonDex~\cite{nondex}.
\Cref{sec:related} explains how NonDex works. This section compares the
errors reported by NonDex and \theDeterminismChecker.

@@

We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
\bugHashSet and \bugClasspath have corresponding test cases whereas the remainder of the bugs do not.
As NonDex is a \textit{flaky test} detection tool, detecting \bugTimestampOutput and \bugTimestampOutput 
is outside of its scope.
The reason for nondeterminism in \bugClasspath was a call to the \<System.getProperty()> method which
is not modeled by NonDex.

@@

\begin{figure}
    \begin{tabular}{lll}
        \<Class> & : \<getDeclaredFields>, \<getDeclaredMethods>, \<getFields> \\
        \<DateFormatSymbols> & : \<getZoneStrings> \\
        \<HashMap> & : \<entrySet>, \<keySet>, \<values>
    \end{tabular}
\caption{Souces of flakiness known to NonDex~\cite{nondex}.}
\label{fig:flaky-sources}
\end{figure}

\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
This section determines whether NonDex finds errors that \theDeterminismChecker does not.
Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
21 of them~\cite{nondex}.
The authors also reported the sources of flakiness
after manually inspecting these tests. 
The flakiness that NonDex found was due to 7 methods in 3 classes (\cref{fig:flaky-sources}).

@@

\section{Comparison to Deflaker}
Deflaker~\cite{deflaker}, like NonDex, reports tests that could be flaky. 
We were unable to run DeFlaker on Randoop, because Deflaker works with
projects built with Maven, but Randoop
uses Gradle for its build system.
%% Don't make promises: they give reviewers an excuse to reject the paper
%% as not a completed research project.
% We plan to convert Randoop's build system to Maven and perform this case study.
%Ideally, we would have 
%liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
%that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
%DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
%in its repository.
%\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
%Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
%\todo{Deflaker requires Maven build. Randoop is on Gradle.}

\subsection{\TheDeterminismChecker on DeFlaker benchmarks}\label{sec:deflaker-benchmarks}
DeFlaker found 87 previously unknown flaky tests in 93 projects that were being actively developed at the time
the paper was written. The authors reported 19 of these tests, out of which 7 were addressed by the maintainers of
those projects~\cite{deflaker}.
We ran \theDeterminismChecker on the part of each of these
codebases where the reported bug was fixed, as in \cref{sec:nondex-benchmarks}.
% The DeFlaker authors graciously fixed problems we discovered while using
% their tool.

>>>>

The state of the art in flaky test detection is NonDex~\cite{ShiGLM2016}.
\Cref{sec:related} explains how NonDex works. This section compares the
errors reported by NonDex and \theDeterminismChecker.

@@

We investigated the tests covering code that \theDeterminismChecker reported as nondeterministic.
\bugHashSet and \bugClasspath have corresponding test cases whereas the remainder of the bugs do not.
As NonDex is a \textit{flaky test} detection tool, detecting bugs that don't have corresponding tests 
is outside of its scope.
The reason for nondeterminism in \bugClasspath was a call to the \<System.getProperty()> method which
is not modeled by NonDex.

@@

\begin{figure}
    \begin{tabular}{lll}
        \<Class> & : \<getDeclaredFields>, \<getDeclaredMethods>, \<getFields> \\
        \<DateFormatSymbols> & : \<getZoneStrings> \\
        \<HashMap> & : \<entrySet>, \<keySet>, \<values>
    \end{tabular}
\caption{Souces of flakiness known to NonDex~\cite{ShiGLM2016}.}
\label{fig:flaky-sources}
\end{figure}

\Cref{sec:nondex-randoop} shows that \theDeterminismChecker finds errors that NonDex does not.
This section determines whether NonDex finds errors that \theDeterminismChecker does not.
Its authors ran NonDex on 195 open-source projects, and NonDex found flaky tests in
21 of them~\cite{ShiGLM2016}.
The authors also reported the sources of flakiness
after manually inspecting these tests. 
The flakiness that NonDex found was due to 7 methods in 3 classes (\cref{fig:flaky-sources}).

@@

\section{Comparison to Deflaker}
Deflaker~\cite{BellLHEYM2018}, like NonDex, reports tests that could be flaky. 
We were unable to run DeFlaker on Randoop, because Deflaker works with
projects built with Maven, but Randoop
uses Gradle for its build system.
%% Don't make promises: they give reviewers an excuse to reject the paper
%% as not a completed research project.
% We plan to convert Randoop's build system to Maven and perform this case study.
%Ideally, we would have 
%liked to run all three tools (NonDex, DeFlaker, and \TheDeterminismChecker) on an open source project
%that wasn't explored by any of the tool for a fair comparison. Unfortunately, there are issues with the
%DeFlaker plugin and we weren't able to run it on any project. We have created a Github issue for the same
%in its repository.
%\todo{Deflaker works now. Tried on retrofit: only works where there is sufficient change in code coverage between head and its parent.
%Looked throught= about 10 commits, didn't find differential code coverage and therefore no flakes}
%\todo{Deflaker requires Maven build. Randoop is on Gradle.}

\subsection{\TheDeterminismChecker on DeFlaker benchmarks}\label{sec:deflaker-benchmarks}
DeFlaker found 87 previously unknown flaky tests in 93 projects that were being actively developed at the time
the paper was written. The authors reported 19 of these tests, out of which 7 were addressed by the maintainers of
those projects~\cite{BellLHEYM2018}.
We ran \theDeterminismChecker on the part of each of these
codebases where the reported bug was fixed, as in \cref{sec:nondex-benchmarks}.
% The DeFlaker authors graciously fixed problems we discovered while using
% their tool.

### 2019-06-05 introduction.tex
edits
Use external bibliography when possible

A nondeterministic program may produce different output on different runs
when provided with the same input.
Nondeterminism is a serious problem for software developers and users.
\begin{itemize}
\item
  Nondeterminism makes a program difficult to \textbf{test}, because test
  oracles must account for all possible behaviors while still enforcing
  correct behaviors.  Test oracles that are too strict lead to flaky
% ,Gyori:2015:RTD:2771783.2771793
  tests~\cite{LuoHEM2014,nondex,deflaker,Rahman:2018:IFF:3236024.3275529,Sudarshan}
  that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
  developers ignore them; in either case, their utility to detect defects
  is limited.
\item
  Nondeterminism makes it difficult to \textbf{compare} two runs of a
  program on different data, or to compare a run of a slightly modified
  program to an original program.  This hinders debugging and maintenance,
  and prevents use of techniques such as Delta Debugging~\cite{Zeller1999,YuLCZ2012}.
%\item
%  Nondeterminism may make outputs incompatible with previous outputs relied
%  on by users or by other systems~\todo{These do not seem to be supporting
%    the claim about bugs related to incompatibilities between outputs.
%    These seem to be general citations about test flakiness or dependence.
%    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,Gyori:2015:RTD:2771783.2771793,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
%  \todo{Citations about bugs related to this?}
\item
  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output~\cite{Tene,Shah}.
\end{itemize}

@@

The high-level goal of our work is to provide programmers with a tool for
specifying deterministic properties in a program and verifying them
statically.
%
Other researchers have also recognized the importance of the problem of nondeterminism.
Previous work in program analysis for nondeterminism has focused on unsound dynamic
approaches that identify flaky test cases.
NonDex~\cite{nondex} uses a modified JVM that returns different results on different
executions, for a few key JDK methods with loose specifications.  Running a
test suite multiple times can reveal unwarranted dependence on those
methods.
%NonDex~\cite{nondex} is a tool that manually\todo{This is confusing:  the
%  tool can't manually identify.  Do you mean that the developers manually
%  identified, then the tool uses those models?}
%identifies 
DeFlaker~\cite{deflaker} looks at a range of commit versions
of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
have been able to identify issues in real-world programs, some of which
have been fixed by the developers. Identifying and
resolving nondeterminism
earlier in the software development lifecycle is beneficial to
developers, because they can avoid bugs associated with flaky tests---reducing
costs~\cite{briski2008minimizing}.

Our analysis uses three main abstractions:
%, or approximations to run-time values:
% MK: the OOPSLA audience knows what an abstraction is
\begin{description}
\item[\<NonDet>] represents values that might differ from run to run.
\item[\<OrderNonDet>] represents collections that are guaranteed to contain
  the same elements but whose iteration order is non-deterministic.
\item[\<Det>] represents values that will be the same across executions.
\end{description}
\noindent
Programmers can write these to specify their program's behavior.
Our full analysis contains other features that increase
expressiveness, permitting it to scale to programs of practical interest.

>>>>

A nondeterministic program may produce different output on different runs
when provided with the same input.
Nondeterminism is a serious problem for software developers and users.
\begin{itemize}
\item
  Nondeterminism makes a program difficult to \textbf{test}, because test
  oracles must account for all possible behaviors while still enforcing
  correct behaviors.  Test oracles that are too strict lead to flaky
% ,GyoriSHM2015
  tests~\cite{LuoHEM2014,ShiGLM2016,BellLHEYM2018,Rahman:2018:IFF:3236024.3275529,Sudarshan}
  that sometimes pass and sometimes fail.  Flaky tests must be re-run, or
  developers ignore them; in either case, their utility to detect defects
  is limited.
\item
  Nondeterminism makes it difficult to \textbf{compare} two runs of a
  program on different data, or to compare a run of a slightly modified
  program to an original program.  This hinders debugging and maintenance,
  and prevents use of techniques such as Delta Debugging~\cite{Zeller1999,YuLCZ2012}.
%\item
%  Nondeterminism may make outputs incompatible with previous outputs relied
%  on by users or by other systems~\todo{These do not seem to be supporting
%    the claim about bugs related to incompatibilities between outputs.
%    These seem to be general citations about test flakiness or dependence.
%    Readers will be unhappy if we make a claim but then do not support it.}\cite{Herzig:2015:EDF:2819009.2819018,Huo:2014:IOQ:2635868.2635917,ZhangJWMLEN2014,BellKMD2015,Fowler,GyoriSHM2015,Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456,Sudarshan}.
%  \todo{Citations about bugs related to this?}
\item
  Nondeterminism reduces users' and developers' \textbf{trust} in a program's output~\cite{Tene,Shah}.
\end{itemize}

@@

The high-level goal of our work is to provide programmers with a tool for
specifying deterministic properties in a program and verifying them
statically.
%
Other researchers have also recognized the importance of the problem of nondeterminism.
Previous work in program analysis for nondeterminism has focused on unsound dynamic
approaches that identify flaky test cases.
NonDex~\cite{ShiGLM2016} uses a modified JVM that returns different results on different
executions, for a few key JDK methods with loose specifications.  Running a
test suite multiple times can reveal unwarranted dependence on those
methods.
%NonDex~\cite{ShiGLM2016} is a tool that manually\todo{This is confusing:  the
%  tool can't manually identify.  Do you mean that the developers manually
%  identified, then the tool uses those models?}
%identifies 
DeFlaker~\cite{BellLHEYM2018} looks at a range of commit versions
of a code, and marks a test as flaky if it doesn't execute any modified code but still fails in the newer version. These techniques
have been able to identify issues in real-world programs, some of which
have been fixed by the developers. Identifying and
resolving nondeterminism
earlier in the software development lifecycle is beneficial to
developers, because they can avoid bugs associated with flaky tests---reducing
costs~\cite{briski2008minimizing}.

Our analysis uses three main abstractions:
%, or approximations to run-time values:
% MK: the OOPSLA audience knows what an abstraction is
\begin{description}
\item[\<NonDet>] represents values that might differ from run to run.
\item[\<OrderNonDet>] represents collections that are guaranteed to contain
  the same elements but whose iteration order is nondeterministic.
\item[\<Det>] represents values that will be the same across executions.
\end{description}
\noindent
Programmers can write these to specify their program's behavior.
Our full analysis contains other features that increase
expressiveness, permitting it to scale to programs of practical interest.

### 2019-06-05 related.tex
Use external bibliography when possible

The state of the art for detecting nondeterministic tests is NonDex~\cite{nondex}.  It is a highly effective tool.
% https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
NonDex uses a hand-crafted list of 47 methods (25 unique method names)
in 13 classes as potential sources of flakiness. 
For each of the identified methods, the authors built models that
return different results when called consecutively. A modified JVM then
runs a given test multiple times and reports the test as being flaky if it observes
diverging test output. While this approach produces precise results, it requires manual inspection
and considerable debugging effort to locate and fix the source of flakiness. \TheDeterminismChecker, in
contrast, reports the cause of nondeterminism at compile time requiring
little to no debugging effort.  However, \theDeterminismChecker requires
much more upfront effort, and it produces false positive warnings.
NonDex's approach of identifying and modeling methods with nondeterministic
specifications is analogous to our library specifications. So far, we have annotated
928 methods across 41 classes in the JDK, in addition to other libraries.
%\todo{Check whether our JDK is missing anything they support - No}
%\todo{Did the NonDex authors contact the maintainers of the tools in which
%  they found nondeterminism - No!}
%\todo{The NonDex authors didn't contact the maintainers of tools.
%From the paper: "NonDex detected 57 previously unknown
%flaky tests in open-source projects and three flaky tests
%that have been already fixed by the open-source software
%developers."}

DeFlaker~\cite{deflaker} is another approach for flaky test detection. It
relies on a
version control history. It computes a diff of the code covered
in the current version and the previous one. If there exists a test case whose code coverage does not include
this diff but still produces different results on the two versions being compared, it is flagged as being flaky.
This approach doesn't require JVM modifications
and integrates easily with production software. DeFlaker reported 19 previously unknown bugs
in open source projects, 7 of which were addressed by the developers of these projects. DeFlaker is agnostic
to the code under test and can therefore report flakiness arising out of concurrency, which \theDeterminismChecker cannot.

Nondeterminism in tests is of interest to both
researchers and software developers alike~\cite{Fowler,Sudarshan}.
Luo et al.~\cite{LuoHEM2014} studied 51 open source projects and analyzed the most common root causes of
test flakiness. The findings suggest that most of the flakiness in tests is caused by
the use of async await, concurrency, or due to test order dependency. The authors of~\cite{Plotkin:1993:LPP:645891.671433}
perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
Additionally,~\cite{Plotkin:1993:LPP:645891.671433} studies the effects of removing these test smells. The results suggest that
refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
these techniques and aims to prevent nondeterminism from causing harmful effects.

@@

Failing tests that are unrelated to code changes can be expensive both in terms of monetary costs and
developer effort. \cite{Herzig:2015:EDF:2819009.2819018} proposes techniques to classify tests as false alarms
if they are known to be caused by testing infrastructure or other environment issues.~\cite{Huo:2014:IOQ:2635868.2635917}
presents an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigates the
effects of the test independence assumption
on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,Gyori:2015:RTD:2771783.2771793}
analyze test dependencies and either prevent them or use this information for other optimizations.
The approaches in \cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
from those caused by source code. While we haven't fully accounted for all nondeterministic sources due
to the environment, our technique is easily extensible by adding annotations to the corresponding classes.

>>>>

The state of the art for detecting nondeterministic tests is
NonDex~\cite{ShiGLM2016}.  It is a highly effective tool.
% https://github.com/TestingResearchIllinois/NonDex/wiki/Supported-APIs
NonDex uses a hand-crafted list of 47 methods (25 unique method names)
in 13 classes as potential sources of flakiness. 
For each of the identified methods, the authors built models that
return different results when called consecutively. A modified JVM then
runs a given test multiple times and reports the test as being flaky if it observes
diverging test output. While this approach produces precise results, it requires manual inspection
and considerable debugging effort to locate and fix the source of flakiness. \TheDeterminismChecker, in
contrast, reports the cause of nondeterminism at compile time requiring
little to no debugging effort.  However, \theDeterminismChecker requires
much more upfront effort, and it produces false positive warnings.
NonDex's approach of identifying and modeling methods with nondeterministic
specifications is analogous to our library specifications. So far, we have annotated
928 methods across 41 classes in the JDK, in addition to other libraries.
%\todo{Check whether our JDK is missing anything they support - No}
%\todo{Did the NonDex authors contact the maintainers of the tools in which
%  they found nondeterminism - No!}
%\todo{The NonDex authors didn't contact the maintainers of tools.
%From the paper: "NonDex detected 57 previously unknown
%flaky tests in open-source projects and three flaky tests
%that have been already fixed by the open-source software
%developers."}

DeFlaker~\cite{BellLHEYM2018} is another approach for flaky test detection. It
relies on a
version control history. It computes a diff of the code covered
in the current version and the previous one. If there exists a test case whose code coverage does not include
this diff but still produces different results on the two versions being compared, it is flagged as being flaky.
This approach doesn't require JVM modifications
and integrates easily with production software. DeFlaker reported 19 previously unknown bugs
in open source projects, 7 of which were addressed by the developers of these projects. DeFlaker is agnostic
to the code under test and can therefore report flakiness arising out of concurrency, which \theDeterminismChecker cannot.

Nondeterminism in tests is of interest to both
researchers and software developers alike~\cite{Fowler,Sudarshan}.
Luo et al.~\cite{LuoHEM2014} studied 51 open source projects and analyzed the most common root causes of
test flakiness. The findings suggest that most of the flakiness in tests is caused by
the use of async await, concurrency, or due to test order dependency. The authors of~\cite{PlotkinA1993}
perform an empirical study on open source benchmarks, the results of which concur with those of~\cite{LuoHEM2014}.
Additionally,~\cite{PlotkinA1993} studies the effects of removing these test smells. The results suggest that
refactoring the test code smells fixed more than half of the identified flaky tests. Our approach is complementary to
these techniques and aims to prevent nondeterminism from causing harmful effects.

@@

Failing tests that are unrelated to code changes can be expensive both in terms of monetary costs and
developer effort. \cite{Herzig:2015:EDF:2819009.2819018} proposes techniques to classify tests as false alarms
if they are known to be caused by testing infrastructure or other environment issues.~\cite{Huo:2014:IOQ:2635868.2635917}
presents an approach that detects brittle assertions in tests by performing a taint analysis on inputs classified as 
controlled and uncontrolled.~\cite{ZhangJWMLEN2014} investigates the
effects of the test independence assumption
on other techniques such as test prioritization, selection, etc. Other approaches~\cite{BellKMD2015,GyoriSHM2015}
analyze test dependencies and either prevent them or use this information for other optimizations.
The approaches in \cite{Dan:2013:10.1007/978-3-642-39038-8_25,Vahabzadeh:2015:7332456} focus on differentiating bugs due to tests
from those caused by source code. While we haven't fully accounted for all nondeterministic sources due
to the environment, our technique is easily extensible by adding annotations to the corresponding classes.
